---
title: "Demo optimizing parameters"
author: "Luis M. Valente, Albert B. Phillimore, Rampal S. Etienne"
date: "19 May 2015"
output: rmarkdown::html_vignette
vignette: >
 %\VignetteIndexEntry{Demo optimizing parameters}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
 collapse = TRUE,
 comment = "#>"
)
```

# DAISIE – Dynamic Assembly of Island biotas through Speciation, Immigration and Extinction 

Citation: Valente LM, Phillimore AB, Etienne RS (2015) Equilibrium and non-
equilibrium dynamics simultaneously operate in the Galápagos islands. Ecology 
Letters, In press. 

## 3.2 Data table

To load the package:

```{r}
library(DAISIE)
```

The raw dataset is inputted as a table. The Galápagos dataset table can be visualized with: 

```{r}
data(Galapagos_datatable)
knitr::kable(Galapagos_datatable)
```

Each row in the table represents an independent colonisation event. 
The table has four columns:

 * `Clade_name`: name of the independent colonization event. 
 * `Status`: One of the following categories: 
    * `Endemic`: applicable for both anagenetic species and radiations. 
    * `Non_endemic`: If the taxon is not endemic to the island, 
      and the age of colonisation is 
      based on a phylogeny where both island and non-island 
      populations of the species 
      have been sampled. 
    * `Non_endemic_MaxAge`: If the taxon is not endemic to the island, 
      and only an upper bound to the time of colonisation 
      of the island is known. This applies if individuals
      from the island population of the species have not been sampled, 
      but an age of the species is known. 
    * `Endemic&Non_Endemic`: When an endemic clade is present 
      and the mainland ancestor has re-colonized. 
      For remote islands this is expected to be very rare.
 * `Missing_species`: Number of island species that 
   were not sampled for a particular clade (only applicable for radiations). 
 * `Branching_times` – This should be the stem age of the 
   population/species in the case of `Non-endemic`, `Non-endemic_MaxAge` 
   and `Endemic` anagenetic species. For cladogenetic 
   species these should be branching times of the radiation 
   including the stem age of the 
   radiation. Note – if there are species within the radiation that 
   are not found on the island 
   (e.g. back-colonisation) the branching times of these species 
   should be excluded, as the 
   mainland species pool is treated as static. 

The same data can also be visualized:

```{r fig.width=7, fig.height=7}
DAISIE::DAISIE_plot_island(Galapagos_datatable, island_age = 4)
```

## 3.3 Formatting table to run in DAISIE 
 
Before running analyses, the datatable needs to be converted to a DAISIE datalist format 
using the function DAISIE_dataprep. 
 
We will prepare two different datalists based on the Galápagos datatable. In the 1st datalist 
we will treat all taxa as equivalent. We will specify an island age of four million years 
(island_age=4) and a mainland pool size of 1000 (M=1000). 

```{r}
data(Galapagos_datatable) 
      
Galapagos_datalist <- DAISIE_dataprep( 
  datatable = Galapagos_datatable, 
  island_age = 4, 
  M = 1000
)
```
 
In the 2nd datalist we will allow for the Darwin’s finches to form a separate group for which 
rates can be decoupled from those governing the macroevolutionary process in all other 
clades (number_clade_types=2 and list_type2_clades = "Finches"). We will set the proportion 
of Darwin’s finch type species in the mainland pool to be 0.163. (prop_type2_pool=0.163). If 
prop_type2_pool is not specified then by default it is given the value of the proportion of 
the Galapagos lineages that Darwin’s finches represent (1/8=0.125 in this case). 

```{r}
data(Galapagos_datatable) 

Galapagos_datalist_2types <- DAISIE_dataprep( 
  datatable = Galapagos_datatable, 
  island_age = 4, 
  M = 1000, 
  number_clade_types = 2, 
  list_type2_clades = "Finches", 
  prop_type2_pool = 0.163
)
```

The objects `Galapagos_datalist` and `Galapagos_datalist_2types` 
can now be run directly in `DAISIE` functions. 

## 3.4 Optimizing parameters using maximum likelihood 
 
The function that conducts maximum likelihood optimization of 
DAISIE model parameters is 
called `DAISIE_ML`. 
 
Different models can be specified using ddmodel option in `DAISIE_ML`: 

 * `ddmodel = 0` : no diversity-dependence 
 * `ddmodel = 1` : linear diversity-dependence in speciation rate 
 * `ddmodel = 11`: linear diversity-dependence in speciation and immigration rate 
 * `ddmodel = 2` : exponential diversity-dependence in speciation rate 
 * `ddmodel = 21`: exponential diversity-dependence in speciation and immigration rate 
 
Different types of parameters can be optimized or fixed. The parameters are given in the 
following order: (1) cladogenesis rate, (2) extinction rate, (3) K’ or carrying capacity 
(maximum number of species that a clade can attain within the island), (4) colonisation rate, 
and (5) anagenesis rate. 
 
The identities of the parameters to be optimized or fixed are specified 
with `idparsopt` and `idparsfix` within the DAISIE_ML function. 
For example, to optimize all parameters we set 
`idparsopt=1:5` and `idparsfix=NULL`. 
To optimize all parameters but fix the rate of extinction, 
we set `idparsopt=c(1,3,4,5)` and `idparsfix=2`. 
To optimize all parameters except cladogenesis 
and anagenesis we set `idparsopt=c(2,3,4)` and `idparsfix=c(1,5)`. 

The values of the parameters to be used as initial values for the optimization are specified 
with `initparsopt`, and the values to be fixed are specified with `parsfix.` For example, if we 
want to optimize all parameters with a starting value of 2 
we set `initparsopt=c(2,2,2,2,2)` and 
`parsfix=NULL`. If we want all starting rates to be 0.1, but K’ to be fixed at 20, we use 
`initparsopt=c(0.1,0.1,0.1,0.1)` and `parsfix=20`. 
 
When running your own data, we strongly recommend that you test multiple initial starting 
parameters for each model, particularly when optimizing models with multiple free 
parameters, as there is a high risk of being trapped in local likelihood sub-optima. We also 
suggest running two rounds of optimization using the optimized parameter set of the 1st 
round as the initial starting values for the 2nd round. Also note that the initial starting values 
in the examples of this tutorial may not be appropriate for your data. 

### Example 1: Optimizing all parameters, with diversity-dependence in speciation and colonisation 
 
We will now optimize all five parameters for a datalist where all clades share the same 
parameters. We will set the model with linear diversity-dependence in speciation rate and in 
immigration rate using ddmodel=11. We will set an initial rate of cladogenesis of 2.5, an 
initial rate of extinction of 2.7, an initial K’ value of 20, an initial rate of colonisation of 0.009 
and an initial rate of anagenesis of 1.01 (`initparsopt = c(2.5,2.7,20,0.009,1.01)`). We will 
optimize all 5 parameters (`idparsopt = 1:5`) and we will fix no parameters (`parsfix = NULL`, 
`idparsfix = NULL`). 
 
```
data(Galapagos_datalist) 

DAISIE_ML( 
  datalist = Galapagos_datalist, 
  initparsopt = c(2.5,2.7,20,0.009,1.01), 
  ddmodel = 11, 
  idparsopt = 1:5, 
  parsfix = NULL, 
  idparsfix = NULL
) 
```


This will take several minutes to run. The parameters optimized and fixed as well as the 
loglikelihood of the initial starting parameters we have set are shown at the top of the 
screen output of DAISIE_ML. Once the optimization is completed, the program will output 
the maximum likelihood parameter estimates and the maximum loglikelihood value. For a 
given dataset, the likelihood of different DAISIE models can be compared with information 
criteria such as BIC and AIC. 

### Example 2: Optimizing model without diversity-dependence 
 
To optimize the parameters of a model with no diversity-dependence, we use the default 
model (ddmodel=0), and fix the parameter number 3 which corresponds to K’ to infinity (Inf).

```

data(Galapagos_datalist) 

DAISIE_ML( 
  datalist = Galapagos_datalist, 
  initparsopt = c(2.5,2.7,0.009,1.01), 
  idparsopt = c(1,2,4,5), 
  parsfix = Inf, 
  idparsfix = 3
) 
``` 
 
### Example 3: Optimizing model with no diversity-dependence and no anagenesis 
 
To optimize the parameters of a model with no diversity-dependence and no anagenesis, we 
use the default model (ddmodel=0), and fix parameters number 3 and 5, which correspond, 
respectively to K’ and rate of anagenesis. 

```
data(Galapagos_datalist) 
DAISIE_ML( 
  datalist=Galapagos_datalist, 
  initparsopt = c(2.5,2.7,0.009), 
  idparsopt = c(1,2,4), 
  parsfix = c(Inf,0), 
  idparsfix = c(3,5)
) 
``` 
 
### Example 4: Optimizing all parameters, but allowing Darwin’s finches to have a separate rate of cladogenesis. 
 
For this example we will use the datalist with Darwin’s finches specified to be of a separate 
type: Galapagos_datalist_2types. 
 
If two types of species are considered, then the parameters of the second type of species are 
in the same order as the first set of parameters, but start at number 6: (6) cladogenesis rate 
of type 2 species, (7) extinction rate of type 2 species, (8) K’ of type 2 species, (9) 
colonisation rate of type 2 species, and (10) anagenesis rate of type 2 species. There is also 
an additional parameter when 2 types of species are considered: the proportion of species 
of type 2 in the mainland pool. This is parameter number 11. 
 
Here we will optimize all parameters, but allow the finches to have a separate rate of 
cladogenesis. We will fix the proportion of type 2 species in the mainland pool at 0.163 
(therefore fixing parameter 11 with idparsfix=11 and parsfix=0.163). Note that because we 
are only allowing the rate of cladogenesis of Darwin’s finches to vary from the background 
rate, we need to specify that the other rates for Darwin’s finches remain the same as the 
background – using idparsnoshift = c(7,8,9,10)). 
 
```
data(Galapagos_datalist_2types) 

DAISIE_ML( 
  ddmodel=11, 
  datalist=Galapagos_datalist_2types, 
  initparsopt= c(0.38,0.55,20,0.004,1.1,2.28), 
  idparsopt = c(1,2,3,4,5,6), 
  parsfix = 0.163, 
  idparsfix = c(11), 
  idparsnoshift = c(7,8,9,10)
) 
```

### Example 5: Optimizing a model with no diversity-dependence, but allowing Darwin’s finches to have a separate rate of cladogenesis and extinction. 

```
data(Galapagos_datalist_2types) 
 
DAISIE_ML( 
  ddmodel=0, 
  datalist=Galapagos_datalist_2types, 
  initparsopt = c(0.38,0.55,0.004,1.1,2.28,2), 
  idparsopt = c(1,2,4,5,6,7), 
  parsfix = c(Inf,0.163), 
  idparsfix = c(3,11), 
  idparsnoshift = c(8,9,10)
) 
``` 
