---
title: "Demo simulating islands"
author: "Luis M. Valente, Albert B. Phillimore, Rampal S. Etienne"
date: "19 May 2015"
output: rmarkdown::html_vignette
vignette: >
 %\VignetteIndexEntry{Demo simulating islands}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
 collapse = TRUE,
 comment = "#>"
)
```

# DAISIE – Dynamic Assembly of Island biotas through Speciation, Immigration and Extinction 
                   
Citation: Valente LM, Phillimore AB, Etienne RS (2015) Equilibrium and non-
equilibrium dynamics simultaneously operate in the Galápagos islands. Ecology 
Letters, In press. 

## 3.2 Data table

To load the package:

```{r}
library(DAISIE)
```

##  3.5 Simulating islands 
 
The function `DAISIE_sim` allows simulation of DAISIE models and plots the results. The user 
specifies the parameters to be simulated, the number of replicates, the length of the 
simulation (typically the island age), and the number of species in the mainland pool. 
 
When the plot_sims option is set to the default (TRUE) the function will produce a species-
through-time plot showing the accumulation of total, endemic and non-endemic species 
through time, as well as confidence intervals for the total number of species. 

```{r}
n_mainland_species <- 1000
island_age <- 4
```

To shorten the run-time of this vignette, reduce the number of `n_replicates`.
For increased accuracy, increase this number.

```{r}
n_replicates <- 10
```
 
### Example 5.1 – Simulating `r n_replicates` islands with no diversity-dependence, all species sharing the  same parameters, and plotting the results 


```{r}
set.seed(42)
```

```{r fig.width=7, fig.height=7}
clado_rate <- 2.550687345 # cladogenesis rate
ext_rate <- 2.683454548 # extinction rate
clade_carr_cap <- Inf # clade-level carrying capacity
imm_rate <- 0.00933207 # immigration rate
ana_rate <- 1.010073119 # anagenesis rate

island_replicates <- DAISIE_sim( 
  time = island_age,
  M = n_mainland_species, 
  pars = c(clado_rate, ext_rate, clade_carr_cap, imm_rate, ana_rate),
  replicates = n_replicates,
  plot_sims = FALSE,
  verbose = FALSE,
  Apars = NULL
)
DAISIE_plot_sims(island_replicates = island_replicates, use_dev_new = FALSE)
```

The object `island_replicates` 
contains the results of the simulation in DAISIE format. 
`r n_replicates` islands are stored in the object, 
and each island replicate can be viewed separately. 
For example, type `island_replicates[[1]]` to view the first replicate.

The element of the list relating to each island contains a table 
with the number of species through time, as well as branching time information for each 
independent colonisation event extant at the end of the simulation.

### Example 5.2 – Simulating `r n_replicates` islands with diversity-dependence (K’=10), all species sharing the same parameters, and plotting the results 

```{r fig.width=7, fig.height=7} 
clado_rate <- 2.550687345 # cladogenesis rate
ext_rate <- 2.683454548 # extinction rate
clade_carr_cap <- 10.0  # clade-level carrying capacity
imm_rate <- 0.00933207 # immigration rate
ana_rate <- 1.010073119 # anagenesis rate

island_replicates_K <- DAISIE_sim( 
  time = island_age, 
  M = n_mainland_species, 
  pars = c(clado_rate, ext_rate, clade_carr_cap, imm_rate, ana_rate),
  replicates = n_replicates,
  plot_sims = FALSE,
  verbose = FALSE
) 
DAISIE_plot_sims(island_replicates_K, use_dev_new = FALSE)
``` 
 
### Example 5.3 – Simulating `r n_replicates` islands allowing Darwin’s finches to have a higher rate of cladogenesis: 

```{r fig.width=7, fig.height=7}
clado_rate_1 <- 0.38 # cladogenesis rate
ext_rate_1 <- 0.55 # extinction rate
clade_carr_cap_1 <- Inf  # clade-level carrying capacity
imm_rate_1 <- 0.04 # immigration rate
ana_rate_1 <- 1.10 # anagenesis rate

clado_rate_2 <- 0.38 # cladogenesis rate
ext_rate_2 <- ext_rate_1 # extinction rate
clade_carr_cap_2 <- clade_carr_cap_1  # clade-level carrying capacity
imm_rate_2 <- imm_rate_1 # immigration rate
ana_rate_2 <- ana_rate_1 # anagenesis rate

island_replicates_2types <- DAISIE_sim( 
  time = island_age,
  M = n_mainland_species, 
  pars = c(
    clado_rate_1, ext_rate_1, clade_carr_cap_1, imm_rate_1, ana_rate_1,
    clado_rate_2, ext_rate_2, clade_carr_cap_2, imm_rate_2, ana_rate_2
  ),
  replicates = n_replicates, 
  prop_type2_pool = 0.163,
  plot_sims = FALSE,
  verbose = FALSE
)
DAISIE_plot_sims(island_replicates_2types, use_dev_new = FALSE)
```                

This produces a figure similar to Fig. 2, with three plots: 
one for the total number of species,  one for species of type 1 and one for species of type 2. Accessing each island replicate 
 individually (e.g. island_replicates_2types[[15]]) shows information on branching times and 
 species-through-time tables for total, type 1 species and type 2 species.